---
title: "01 - File Formats"
format: html
editor: visual
---

## 🧬 Common Bioinformatics File Formats

In bioinformatics, different types of biological data are stored in specialized file formats. These include:

-   Sequence data (e.g., DNA, RNA, proteins)
-   Quality scores from sequencing machines
-   Annotations, features, and metadata

In this notebook, you’ll learn how to **identify**, **view**, and **read** several common formats using R packages from Bioconductor.

All file examples are stored in the `../data/` folder.

### 📘 FASTA Format

FASTA is one of the most common formats for storing nucleotide or protein sequences. Each record starts with a `>` followed by a sequence ID and optional description, then one or more lines of sequence.

```{r}
library(Biostrings)
fasta_path <- "data/fasta_example.fasta"
fasta_seqs <- readDNAStringSet(fasta_path)

# Print ID and sequence
for (i in seq_along(fasta_seqs)) {
  cat("ID:", names(fasta_seqs)[i], "\n")
  cat("Sequence:", as.character(fasta_seqs[[i]]), "\n\n")
}

```

### 📗 FASTQ Format

FASTQ stores sequencing reads and quality scores. Each record consists of: 1. `@` followed by sequence ID 2. Sequence string 3. `+` separator 4. ASCII-encoded quality scores

```{r}
library(ShortRead)
fastq_path <- "data/fastq_example.fastq"
fastq_reads <- readFastq(fastq_path)

# Print sequence IDs, sequences, and qualities
for (i in 1:length(fastq_reads)) {
  cat("ID:", as.character(id(fastq_reads[i])), "\n")
  cat("Sequence:", as.character(sread(fastq_reads[i])), "\n")
  cat("Quality:", as.character(quality(quality(fastq_reads[i]))), "\n\n")
}

```


### 📙 BED Format

BED is used for representing genomic regions or features. We can import BED files using `rtracklayer`.

```{r}
library(rtracklayer)
bed_path <- "data/bed_example.bed"
bed_data <- import(bed_path)
head(bed_data)

```

### 📒 SAM/BAM Format

These formats store alignments from sequencing reads to reference genomes. We'll use `Rsamtools` to load BAM files.

```{r}
library(Rsamtools)
bam_path <- "data/bam_example.bam"
bam_data <- scanBam(bam_path)
bam_data[[1]][1:3]

```

### 🧾 VCF Format

VCF is used for storing genetic variants. Use `VariantAnnotation` to read VCF files.

```{r}
library(VariantAnnotation)
vcf_path <- "data/vcf_example.vcf"
vcf <- readVcf(vcf_path, "hg19")
vcf

```

## ✅ Summary

-   R can handle a wide range of bioinformatics file formats.
-   Packages from Bioconductor like `Biostrings`, `ShortRead`, `rtracklayer`, `genbankr`, `Rsamtools`, and `VariantAnnotation` make it easy to import and explore biological data.
