---
title: "Basic Sequence Handling"
format: html
editor: visual
---

### Basic Sequence Handling

R provides powerful Bioconductor packages to work with biological sequences, such as `Biostrings` and `seqinr`. In this notebook, we explore how to manipulate DNA, RNA, and protein sequences.

#### ðŸ§¬ Creating and Inspecting a DNA Sequence

We start by creating a DNA sequence and printing basic information.

```{r}
library(Biostrings)
library(seqinr)

# Create DNA sequence
seq <- DNAString("ATGGCCATTGTAATGGGCCGCTGAAAGGGTGCCCGATAG")
cat("Original:", as.character(seq), "\n")
```

#### ðŸ” Complement and Reverse Complement

```{r}
cat("Complement:", as.character(complement(seq)), "\n")
cat("Reverse:", as.character(reverse(seq)), "\n")
cat("Reverse Complement:", as.character(reverseComplement(seq)), "\n")
```

#### ðŸ”  Counting Nucleotides

```{r}
nt_counts <- table(strsplit(as.character(seq), "")[[1]])
print(nt_counts)
```

#### ðŸ”„ Transcription and Translation

```{r}
# Transcription (DNA to RNA)
rna <- RNAString(chartr("T", "U", as.character(seq)))
cat("Transcription:", as.character(rna), "\n")

# Translation
protein <- Biostrings::translate(seq)
cat("Translation:", as.character(protein), "\n")

# Back-transcription
back_dna <- chartr("U", "T", as.character(rna))
cat("Back-transcribed DNA:", back_dna, "\n")
```

#### ðŸ”¢ GC Content Calculation

```{r}
gc_content <- GC(s2c(as.character(seq)))
cat("GC Content:", gc_content, "\n")
```

#### ðŸ§¬ Codon Table & Genetic Code Info

```{r}
# Helper function to format codon table output
printCodonTable <- function(code, label) {
  cat(paste0("\nðŸ§¬ ", label, ":\n"))
  for (i in seq_along(code)) {
    cat(sprintf("%s â†’ %s\n", names(code)[i], code[i]))
  }
}

# Show a few available tables
cat("ðŸ§¬ Available Genetic Code Tables:\n")
print(names(GENETIC_CODE_TABLE))

# Standard Code
std_code <- getGeneticCode()
printCodonTable(std_code[1:10], "Standard Genetic Code (first 10)")

# Vertebrate Mitochondrial
mito_code <- getGeneticCode("2")
printCodonTable(mito_code[1:10], "Vertebrate Mitochondrial Code (first 10)")

# Invertebrate Mitochondrial
inv_code <- getGeneticCode("5")
printCodonTable(inv_code[1:10], "Invertebrate Mitochondrial Code (first 10)")
```

#### ðŸ”Ž Pattern Matching and Motif Search

```{r}
cat("Base frequencies:\n")
print(alphabetFrequency(seq, baseOnly = TRUE))

cat("Subsequence (positions 5 to 15):\n", as.character(subseq(seq, 5, 15)), "\n")

motif <- "ATG"
cat("Motif search ('ATG'):\n")
print(matchPattern(motif, seq))

cat("Motif count ('ATG'):\n")
print(countPattern(motif, seq))
```

#### ðŸš¦ Start and Stop Codons

```{r}
codon_table <- getGeneticCode()
start_codons <- names(codon_table)[codon_table == "M"]
stop_codons <- names(codon_table)[codon_table == "*"]
cat("\nStart codons:", paste(start_codons, collapse = ", "), "\n")
cat("Stop codons:", paste(stop_codons, collapse = ", "), "\n")
```

#### ðŸ¤ Pairwise Alignment

```{r}
seq2 <- DNAString("ATGGCCATTGTAATGGCCTGCTGAAAGGGTGCCCGATAG")
alignment <- pairwiseAlignment(seq, seq2)
cat("Pairwise alignment:\n")
print(alignment)
```

## âœ… Summary

-   R allows powerful sequence operations using `Biostrings` and `seqinr`
-   You can manipulate and analyze DNA, RNA, and proteins
-   Common tasks like GC content, transcription, and translation are supported
