---
title: "NCBI Database Access"
format: html
editor: visual
---

# ðŸ”Ž Accessing NCBI Databases with R

This notebook demonstrates how to use the `rentrez` package to access biological data from **NCBI (National Center for Biotechnology Information)**.

We'll: - List available NCBI databases - Search for the **BRCA1** gene in humans - Fetch GenBank records - Display sequence summaries

## ðŸ“š Step 1: View Available NCBI Databases

```{r}
library(rentrez)
dbs <- entrez_dbs()
cat("Total databases:", length(dbs), "\n")
head(dbs, 10)

```

## ðŸ” Step 2: Search for BRCA1 Gene in Humans (nucleotide db)

```{r}

library(rentrez)

# Optional: set email directly in each function call
results <- entrez_search(db = "nucleotide",
                         term = "BRCA1[Gene Name] AND Homo sapiens[Organism]",
                         retmax = 10,
                         email = "blagoj4e@gmail.com")


cat("Found IDs:", paste(results$ids, collapse=", "), "\n")

```

## ðŸ“¥ Step 3: Fetch GenBank Records

```{r}
if (length(results$ids) > 0) {
  gb_data <- entrez_fetch(db = "nucleotide", 
                          id = results$ids, 
                          rettype = "gb", 
                          retmode = "text")
  # Save to file
  writeLines(gb_data, "brca1_records.gb")
  cat("Saved GenBank records to brca1_records.gb\n")
} else {
  cat("No IDs found.\n")
}

```

## ðŸ“– Step 5: Preview GenBank Text

```{r}
# Show first few lines of downloaded file
readLines("brca1_records.gb", n = 20)

```

## âœ… Summary

-   Used `rentrez` to connect to NCBI
-   Performed gene search
-   Fetched GenBank data
-   Saved and inspected sequence records
