---
title: "BLAST Practical Flow"
format: html
editor: visual
---

# 🧬 Introduction to BLAST in R

**BLAST** (Basic Local Alignment Search Tool) allows you to compare a DNA or protein sequence against a database to find similarities. This notebook demonstrates a BLAST workflow in R using `rentrez`, `genbankr`, and local tools.

We will: - Fetch the *E. coli* genome from NCBI - Extract coding sequences and proteins - Save a protein in FASTA format - Provide instructions to run BLAST (locally or via NCBI)

## 📥 Step 1: Fetch E. coli K12 Genome from NCBI

```{r}
library(rentrez)

gb <- entrez_fetch(db = "nucleotide", id = "U00096.3", rettype = "gb", retmode = "text", email = "blagoj4e@gmail.com")
writeLines(gb, "ecoli_k12.gb")
cat("Genome saved to ecoli_k12.gb\n")

```

## 🧬 Step 2: Parse GenBank and Extract Proteins

```{r}
extract_protein_translations <- function(file_path, min_length = 20) {
  lines <- readLines(file_path)
  
  translations <- c()
  inside_feature <- FALSE
  current_translation <- ""
  
  for (line in lines) {
    # Detect start of a CDS feature
    if (grepl("^\\s{5}CDS\\s", line)) {
      inside_feature <- TRUE
      current_translation <- ""
    }
    
    # Detect translation tag inside a CDS
    if (inside_feature && grepl("/translation=", line)) {
      current_translation <- gsub('.*="/?|"$', '', line)
      # Sometimes translation spans multiple lines
    } else if (inside_feature && current_translation != "" && grepl('^[ ]{21}[A-Z]+', line)) {
      current_translation <- paste0(current_translation, gsub(" ", "", line))
    } else if (inside_feature && current_translation != "" && !grepl('^[ ]{21}[A-Z]+', line)) {
      # End of translation
      inside_feature <- FALSE
      current_translation <- gsub('"', '', current_translation)
      if (nchar(current_translation) > min_length) {
        translations <- c(translations, current_translation)
      }
      current_translation <- ""
    }
  }
  
  cat("✅ Found", length(translations), "proteins with valid translations.\n")
  return(translations)
}

proteins <- extract_protein_translations("ecoli_k12.gb")
head(proteins, 3)  # Show first 3 proteins
```

## 💾 Step 3: Save a Protein to FASTA

```{r}
library(Biostrings)
protein_seqs <- extract_protein_translations("ecoli_k12.gb")
aa <- AAStringSet(protein_seqs[1])
names(aa) <- "ecoli_protein_1"
writeXStringSet(aa, filepath = "ecoli_protein.fasta")
cat("Saved ecoli_protein.fasta\n")

```

```{r}
# Run this after you've saved `ecoli_protein.fasta`

# Run BLAST remotely using system call
system2("blastp",
        args = c("-query", "ecoli_protein.fasta",
                 "-db", "nr",
                 "-remote",
                 "-out", "blast_result.xml",
                 "-outfmt", "5"))

# Now parse and show hits
library(xml2)

blast_xml <- read_xml("blast_result.xml")
hits <- xml_find_all(blast_xml, ".//Hit")

cat("Top hits:\n")
for (i in seq_len(min(3, length(hits)))) {
  hit_id <- xml_text(xml_find_first(hits[[i]], ".//Hit_id"))
  hit_def <- xml_text(xml_find_first(hits[[i]], ".//Hit_def"))
  score <- xml_text(xml_find_first(hits[[i]], ".//Hsp_bit-score"))
  cat(paste0("• ", hit_id, " — ", substr(hit_def, 1, 60), "... [Score: ", score, "]\n"))
}
```

## ✅ Summary

-   Fetched GenBank genome
-   Parsed CDS features and extracted protein
-   Saved a selected protein in FASTA format
-   Provided instructions for BLAST query submission
