---
title: "Data Visualization"
format:
  html:
    toc: true
    embed-resources: true  # Optional: for self-contained file
editor: visual
---

# 📊 Bioinformatics Data Visualization in R

This notebook demonstrates how to visualize: - CDS sequence properties (length, GC content) - Simulated gene expression (volcano plot) - 3D molecular structures (alternative tools)

We’ll use real bacterial CDS and simulate gene data.

## 1️⃣ Download E. coli CDS FASTA

```{r}
url <- "https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/005/845/GCF_000005845.2_ASM584v2/GCF_000005845.2_ASM584v2_cds_from_genomic.fna.gz"
destfile <- "ecoli_cds.fna.gz"
download.file(url, destfile)
cat("✅ Downloaded E. coli CDS FASTA\n")

```

## 2️⃣ Parse FASTA and Compute Length + GC

```{r}
library(Biostrings)

cds <- readDNAStringSet("ecoli_cds.fna.gz")
seq_lengths <- width(cds)

# GC content calculation
gc_content <- letterFrequency(cds, letters = c("G", "C"), as.prob = TRUE)
gc_content_total <- rowSums(gc_content)

```

## 📈 Plot 1: Length Distribution

```{r}
library(ggplot2)
df <- data.frame(Length = seq_lengths)

ggplot(df, aes(x = Length)) +
  geom_histogram(bins = 50, fill = "lightblue") +
  labs(title = "CDS Length Distribution", x = "Length (bp)", y = "Frequency") +
  theme_minimal()

```

## 📉 Plot 2: GC Content vs Length

```{r}
df$GC <- gc_content_total

ggplot(df, aes(x = Length, y = GC)) +
  geom_point(alpha = 0.5) +
  labs(title = "GC Content vs CDS Length", x = "Length (bp)", y = "GC Content") +
  theme_minimal()

```

## 🧪 Simulated Volcano Plot (Gene Expression)

```{r}
set.seed(123)
gene_data <- data.frame(
  log2FoldChange = rnorm(1000, 0, 2),
  negLogP = -log10(runif(1000, 0.0001, 1))
)

ggplot(gene_data, aes(x = log2FoldChange, y = negLogP)) +
  geom_point(alpha = 0.4) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  labs(title = "Simulated Volcano Plot", x = "log2 Fold Change", y = "-log10(p-value)") +
  theme_minimal()

```

## 🧬 Optional: 3D Structure Viewer

To view protein structures interactively:

-   Use <https://www.rcsb.org/3d-view> and upload or search for a PDB file
-   In Quarto HTML, you can embed `r3dmol` widgets (JavaScript)
-   In Shiny, use `r3dmol::r3dmol()` to show structures

## ✅ Summary

-   Downloaded real E. coli CDS
-   Visualized sequence length and GC content
-   Simulated gene expression plot
-   Described 3D structure viewing options \## 🧬 Optional: Interactive 3D Structure with r3dmol

To embed 3D molecular structures in **Quarto HTML documents**, you can use the `{r3dmol}` package.

### Install and Load Package

```{r}
# install.packages("r3dmol")  # Uncomment if not installed
library(r3dmol)
```

### Example: Load PDB 1BNA (DNA Duplex)

```{r}
library(r3dmol)
library(magrittr)
r3dmol() %>%
  m_add_model(data = '1BNA', format = 'pdb') %>%
  m_set_style(style = m_style_cartoon()) %>%
  m_zoom_to()
```

**Note**: This will only render in a Quarto HTML output or a Shiny app, not in static notebooks or PDF.
